Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2020-08-05 write-up/fda-recurrent.tex
updated paper

\subsection{Estimation of the windowed covariate history}

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(t,s) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(t,s) = X(t,s) - \hat \mu_y(t,s)$ to be the mean-zero process at each time~$t \in \bfT_i \cup \bfD_i$.

@@

\subsection{Connection to logistic score functions}

@@

Given each simulated user-day, we then
\begin{itemize}
\item Sample every 0.5 hours; then subsample at each subrate. Estimates are given by $\hat \beta_{i,0.5} (t)$ and $\hat \beta_{i,1} (t)$.
\item Bias is average over the estimates $\bar \hat \beta_{0.5} (t) = 100^{-1} \sum_{i=1}^100 \hat \beta_{i, 0.5} (t)$ compared to truth.
\item Take squared difference integrated to get variance.
\item To get the subsampling variance take $1/1000 \sum_{i=1}^1000 (\hat \beta_{i,0.5} (t) - \hat \beta_{i,1} (t))^2$.
\item Normalize everything by squared function!
\end{itemize}

We sample non-events at rates of 1 per 4/2/1/0.5 hours.

Since we are primarily interested in accuracy of the final estimates~$\hat \beta(s)$, we report the mean integrated squared error (MISE) of~$\hat \beta (s)$, defined as~$\frac{1}{1000} \sum_{j=1}^{1000} \int_{0}^{\Delta^\prime} ( \hat \beta_j (s) - \beta(s))^2 ds$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  We define the MISE in this manner to account for settings where~$\Delta$ is unknown. Table~\ref{tab:mise} show the MISE decomposed into the variance and squared biase

\begin{table}[!th]
\begin{tabular}{c | p{2cm} p{2cm} p{2cm} p{2cm}}
& \multicolumn{4}{c}{Sampling rate} \\ \cline{2-5}
 & 0.5 & 1 & 2 & 4 \\ \hline
Subsampling variance & - & $9.0 \times 10^{-7}$ & $2.9 \times 10^{-6}$ & $5.8 \times 10^{-6}$  \\
Variance & $7.9 \times 10^{-3}$ & $9.2 \times 10^{-3}$ & $1.0 \times 10^{-2}$ & $1.30 \times 10^{-2}$ \\
Squared Bias & $9.8 \times 10^{-2}$ & $9.5 \times 10^{-2}$ & $9.3 \times 10^{-2}$ & $9.2 \times 10^{-2}$   \\
MISE & $1.1 \times 10^{-1}$ & $1.0 \times 10^{-1}$ & $1.0 \times 10^{-1}$ & $1.0 \times 10^{-1}$ \\ \hline
Avg. runtime (secs) & 317 & 177 & 104 & 68 \\ \hline
Subsampling variance & - & $2.9 \times 10^{-6}$ & $8.9 \times 10^{-6}$ & $2.0 \times 10^{-5}$  \\
Variance & $1.0 \times 10^{-2}$ & $1.3 \times 10^{-2}$ & $2.0 \times 10^{-2}$ & $3.1 \times 10^{-2}$ \\
Squared Bias & $6.9 \times 10^{-2}$ & $6.5 \times 10^{-2}$ & $6.0 \times 10^{-2}$ & $5.4 \times 10^{-2}$   \\
MISE & $7.9 \times 10^{-2}$ & $7.9 \times 10^{-2}$ & $8.0 \times 10^{-2}$ & $8.4 \times 10^{-2}$ \\ \hline
Avg. runtime (secs) & 36 & 23 & 15 & 12 \\ \hline
\end{tabular}
\caption{Mean-integrated squared error and type I error as a function of sampling rate for $\beta(s)$ given by (1) and (2) respectively.}
\label{tab:mise}
\end{table}

@@

In this section, we demonstrate the flexibility of our approach by exploring extensions in several important directions to ensure these methods are robust for practical use in mobile health. This section will continue to leverage the connection to generalized functional linear models provided by Lemma~\ref{lemma:logistic}.

\subsection{Multiple sensors}
\label{section:multiplesensors}

\begin{equation}
\label{eq:multihazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_{1i} (s) \beta_1(s) ds + \int_{t-\Delta}^t x_{2i} (s) \beta_2(s) ds  \right)
\end{equation}

\subsection{Missing data}
\label{section:missingdata}
\walt{Question: How to perform multiple imputation for~$X(\cdot,t)$
  and $X(\cdot, t^\prime)$ for~$|t - t^\prime| <\Delta$? That is, how
  do you perform multiple imputation with this overlapping structure?}

A core concern in mHealth is that sensor data can often be missing for intervals of time due to sensor wearing issue (e.g., sensor band becomes loose, turns off accidentally, data packets lost, or battery dies). In the suicidal ideation case study, for example, there are 2139 self-identified moments of distress across all 91 participants. Of these, 1289 event times had complete data for the prior thirty minutes, 1984 had fraction of missing data on a fine grid less than 30\%, and 1998 had fraction of missing data on a fine grid less than 10\%.

Missing data is a critical issue because~$c_{i,k} (t)$ cannot be estimated if~$X(s,t)$ is not observed for all~$s \in [0,\Delta]$. Moreover, we want standard errors to reflect the uncertainty in these coefficients when missing data is prevalent. \cite{Goldsmith2011} suggest using best linear unbiased predictors (BLUP) or posterior modes in the mixed effects model to estimate~$c_{i,k} (t)$; however, one criticism would be when there is substantial variability in these estimates.  To deal with this, \cite{Crainiceanu2010} take a full Bayesian analysis. \cite{Yao2005} introduced PACE as an alternative frequentist method. However,~\cite{Petrovich2018} shows that for sparse, irregular longitudinal, the imputation model should not ignore the outcome variable~$Y_i (t)$.

@@

@@

>>>>

\subsection{Estimation of the windowed covariate history}
\label{sec:margcov}
We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(t,s) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(t,s) = X(t,s) - \hat \mu_y(t,s)$ to be the mean-zero process at each time~$t \in \bfT_i \cup \bfD_i$.

@@

\subsection{Connection to logistic score functions}
\label{eq:logistication}

@@

We generate $1000$ datasets, each consisting of $500$ user-days.  For a given simulated user-day, we randomly sample non-event times using a Poisson process with rate of a halfhour.  We use the proposed methodology to construct the estimate $\hat \beta_{i,0.5} (t)$ for the $i$th simulated dataset; we then subsample the sampled non-event times with thinning probability $1/2$, $1/4$ and $1/8$.  This results in randomly sampled non-event times given by a Poisson process with rates of an hour, two hours, and four hours.  We can construct the corresponding estimates: $\hat \beta_{i,1} (t)$, $\hat \beta_{i,2} (t)$, and $\hat \beta_{i,4} (t)$ respectively. Subsampling allows us to compare the variance due to subsampling as compared to the variance due to sampling fewer non-event times.

Since we are primarily interested in accuracy, we report the mean integrated squared error (MISE) defined as~$\frac{1}{1000} \sum_{i=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \beta(s))^2 ds$ for $j=0.5,1,2,4$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  The MISE is defined in this manner to account for settings where~$\Delta$ is unknown.  Next, let $\bar \beta_{j} (t) = \frac{1}{1000} \sum_{i=1}^{1000} \hat \beta_{i,j} (t)$ denote the average estimate for $j=0.5,1,2,4$.  Then the squared bias is given by $\int_{0}^{\infty} ( \bar \beta_{j} (s) - \beta (s))^2 ds$ and the variance is given by $\frac{1}{1000} \sum_{j=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \bar \beta_j(s))^2 ds$.  The subsampling variance is defined as $\frac{1}{1000} \sum_{i=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \hat \beta_{i,0.5} (s))^2 ds$.  Table~\ref{tab:mise} show the MISE decomposed into the variance and squared biased as well as the subsampling variance.  To allow fair comparisons across the two choices of $\beta(s)$, all reported numbers are scaled by the integrated square of the true function $\int_{0}^{\infty} \beta(s)^2 ds$. The average runtime (in seconds) is also reported.

Table~\ref{tab:mise} demonstrates that the variance does increase as the sampling rate decreases.  However, the rate of increase in the MISE is quite low relative.  In the second case study, for example, the MISE increases by 6\% while the run time is 3 times faster.  In the first case study, the MISE remains roughly constant but the run time is 4.7 times faster.  This highlights the efficiency-computation trade-off.  Extrapolating to a complete data analysis, we can see that the run time will continue to increase significantly with minimal improvement in the mean integrated squared error.

\begin{table}[!th]
\begin{tabular}{c | p{2cm} p{2cm} p{2cm} p{2cm}}
& \multicolumn{4}{c}{Sampling rate} \\ \cline{2-5}
 & 0.5 & 1 & 2 & 4 \\ \hline
Subsampling variance & - & $9.0 \times 10^{-7}$ & $2.9 \times 10^{-6}$ & $5.8 \times 10^{-6}$  \\
Variance & $7.9 \times 10^{-3}$ & $9.2 \times 10^{-3}$ & $1.0 \times 10^{-2}$ & $1.30 \times 10^{-2}$ \\
Squared Bias & $9.8 \times 10^{-2}$ & $9.5 \times 10^{-2}$ & $9.3 \times 10^{-2}$ & $9.2 \times 10^{-2}$   \\
MISE & $1.1 \times 10^{-1}$ & $1.0 \times 10^{-1}$ & $1.0 \times 10^{-1}$ & $1.0 \times 10^{-1}$ \\ \hline
Avg. runtime (secs) & 317 & 177 & 104 & 68 \\ \hline
Subsampling variance & - & $2.9 \times 10^{-6}$ & $8.9 \times 10^{-6}$ & $2.0 \times 10^{-5}$  \\
Variance & $1.0 \times 10^{-2}$ & $1.3 \times 10^{-2}$ & $2.0 \times 10^{-2}$ & $3.1 \times 10^{-2}$ \\
Squared Bias & $6.9 \times 10^{-2}$ & $6.5 \times 10^{-2}$ & $6.0 \times 10^{-2}$ & $5.4 \times 10^{-2}$   \\
MISE & $7.9 \times 10^{-2}$ & $7.9 \times 10^{-2}$ & $8.0 \times 10^{-2}$ & $8.4 \times 10^{-2}$ \\ \hline
Avg. runtime (secs) & 36 & 23 & 15 & 12 \\ \hline
\end{tabular}
\caption{Mean-integrated squared error, varaince, squared bias, and subsampling variance for $\beta(s)$ given by (1) and (2) respectively.}
\label{tab:mise}
\end{table}

@@

In this section, we demonstrate the flexibility of the proposed approach by exploring extensions in several important directions to ensure these methods are robust for practical use with high frequency data. This section will continue to leverage the connection to generalized functional linear models provided by Lemma~\ref{lemma:logistic}.

\subsection{Multivariate extensions}
\label{section:multiplesensors}

In this section, we extend our model to the case of multiple functional regressors.  That is, suppose $L$ health processes, i.e., $\bfx_i = \{ \bfx_i (t) = (x_{i,1} (t), \ldots, x_{i,L} (t)) \}_{0 < s < \tau_i}$, for each participant is measured at a dense grid of time points. In the suicidal ideation case study, for example, accelerometer is measured at a rate of 32Hz while electrodermal activity (EDA) is measured at a rate of 4Hz.  A multivariate extension of our model~\eqref{eq:hazard} is given by
\begin{equation}
\label{eq:multihazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_{i,1} (s) \beta_1(s) ds + \cdots + \int_{t-\Delta}^t x_{i,L} (s) \beta_2(s) ds  \right)
\end{equation}
The approach given in Section~\ref{eq:logistication} extends naturally to the multivariate functional setting. For each functional regressor, we estimate the pooled sample covariance $\Sigma_{y,l}$ for $y \in \{0,1\}$ and $l=1,\ldots,L$ as in Section~\ref{sec:margcov}.  Let $\sum_{k=1}^\infty \hat \lambda^{(y)}_{k,l} \hat \psi^{(y)}_{k,l} (s) \hat \psi_{k,l}(t)$ be the spectral decomposition of $\hat \Sigma_{y,l}$.  Then $x_{i,j}(t)$ is approximated using a truncated Karhunen-Loeve

, the coefficient is expressed using separate marginal covaraiance
$$
\int_0^\Delta x_{il} \beta(t) =
$$

Note that we express each coefficient function in the same spline basis; however, different basis expansions are used for each health process as defined by the marginal covariance calculation in Section~\ref{sec:margcov}.

\subsection{Missing data}
\label{section:missingdata}
% \walt{Question: How to perform multiple imputation for~$X(\cdot,t)$
%   and $X(\cdot, t^\prime)$ for~$|t - t^\prime| <\Delta$? That is, how
%   do you perform multiple imputation with this overlapping structure?}

A core concern in mHealth is that sensor data can often be missing for intervals of time due to sensor wearing issues (e.g., sensor band becomes loose, turns off accidentally, data packets lost, or battery dies). In the suicidal ideation case study, for example, there are 2139 self-identified moments of distress across all 91 participants. Of these, 1289 event times had complete data for the prior thirty minutes, 1984 had fraction of missing data on a fine grid less than 30\%, and 1998 had fraction of missing data on a fine grid less than 10\%.

Missing data is a critical issue because~$c_{i,k} (t)$ cannot be estimated if~$X(s,t)$ is not observed for all~$s \in [0,\Delta]$. Moreover, standard errors should reflect the uncertainty in these coefficients when missing data is prevalent. \cite{Goldsmith2011} suggest using best linear unbiased predictors (BLUP) or posterior modes in the mixed effects model to estimate~$c_{i,k} (t)$; however, one criticism would be when there is substantial variability in these estimates.  To deal with this, \cite{Crainiceanu2010} take a full Bayesian analysis. \cite{Yao2005} introduced PACE as an alternative frequentist method. However,~\cite{Petrovich2018} shows that for sparse, irregular longitudinal, the imputation model should not ignore the outcome variable~$Y_i (t)$.

@@

The above procedure is

@@

Recommendation from~\cite{Bartlett2020}.
